# Reproducible Research: Peer Assessment 1

```{r setup}
# First let's do some setup.
library(knitr)
library(ggplot2)
library(plyr)
library(lattice)
opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Data are loaded from the data directory. The "date" field is processed using `as.Date`.

```{r loaddata}
df <- read.csv("../data/activity.csv")

#Convert dates to Date class.
df$date <- as.Date(df$date)

# Convert time values to strings and add leading zeroes.
df$interval <- sprintf("%04d", df$interval)
```


## What is mean total number of steps taken per day?

First, sum up the steps taken each day, ignoring missing values. Those are presented on a histogram

```{r sumsteps}
total_steps_by_day <- tapply(df$steps, df$date, FUN = sum, na.rm = TRUE)
histogram(total_steps_by_day, breaks = 
             seq(0, max(total_steps_by_day) + 2000, by = 2000), 
     xlab = "Total number of steps in a day", 
     main = "Histogram of steps by day")
```

Calculate the mean number of steps.
```{r}
mean(total_steps_by_day)
```

And the median.
```{r}
median(total_steps_by_day)
```


## What is the average daily activity pattern?

Create a list containing the intervals in character format. Then calculate the mean number of steps taken in those intervals, with missing values ignored. The result is plotted vs. time.


```{r meantime}
interval_val <- unique(df$interval)
mean_steps_by_time <- data.frame(row.names = interval_val, 
                                 steps = as.numeric(
                                         rep(NA, length(interval_val))))
for (x in interval_val) {
        mean_steps_by_time[x,] = 
                mean(subset(df, interval == x)$steps, na.rm = TRUE)
}

plot(x = rownames(mean_steps_by_time), y = mean_steps_by_time[,1], type = "l", xlab = "time interval", ylab = "mean number of steps")
```

The maximum mean steps is at time:

```{r maxtime}
rownames(mean_steps_by_time)[which.max(mean_steps_by_time$steps)]
```

## Imputing missing values

There are this many intervals with missing values:

```{r countna}
sum(is.na(df$steps))
```

We create a duplicate data frame to fill in. If the number of steps during an interval is unknown, the average for that time of day is used.

```{r fillna}
# Duplicate df.
filled_df <- df

# Look up mean steps at that time for NA values.
for (x in which(is.na(filled_df$steps))) {
        filled_df[x,]$steps <- 
                as.integer(mean_steps_by_time[filled_df[x,]$interval,])
}
```

Calcluate again the new total number of steps per day, this time with missing values imputed from the average for that time interval.
```{r}
total_filled_steps_by_day <- 
        tapply(filled_df$steps, filled_df$date, FUN = sum, na.rm = FALSE)
histogram(total_filled_steps_by_day, breaks = 
             seq(0, max(total_filled_steps_by_day) + 2000, by = 2000), 
     xlab = "Total number of steps in a day", 
     main = "Histogram of steps by day, with missing data imputed")
```

Calculate the mean number of steps with the imputed data.
```{r}
mean(total_filled_steps_by_day)
```

And the median.
```{r}
median(total_filled_steps_by_day)
```

Both values are somewhat increased.


## Are there differences in activity patterns between weekdays and weekends?

First, add in the day of the week to the data frame.
```{r}
# Maybe could have done this to begin. Fix later.
weekends <- c("Saturday", "Sunday")
filled_df$weekday <- as.factor(ifelse(weekdays(filled_df$date) %in% weekends, 
                               "Weekend", "Weekday"))
```

First calculate the mean number of steps per time interval, split between weekdays and weekend days. Those values are plotted.

```{r meantimeweekday, fig.height = 8}
mean_steps_by_time_weekday <- data.frame(row.names = interval_val, 
                                 steps_weekday = as.numeric(
                                         rep(NA, length(interval_val))),
                                 steps_weekend = as.numeric(
                                         rep(NA, length(interval_val))))

## Need to clean this up later.
for (x in interval_val) {
        mean_steps_by_time_weekday[x,]$steps_weekday = 
                mean(subset(filled_df, interval == x 
                            & weekday == "Weekday")$steps, na.rm = FALSE)
        mean_steps_by_time_weekday[x,]$steps_weekend = 
                mean(subset(filled_df, interval == x 
                            & weekday == "Weekend")$steps, na.rm = FALSE)    
}

## Create new data frame with weekday and weekend steps in one column, 
## with factor to distinguis them. Need this to use lattice.

mean_steps_by_time_weekday2 <- 
        data.frame(interval = rownames(mean_steps_by_time_weekday), 
                   steps = mean_steps_by_time_weekday$steps_weekday, 
                   day = rep("Weekday", 288))

mean_steps_by_time_weekday3 <- 
        rbind(mean_steps_by_time_weekday2, 
              data.frame(interval = rownames(mean_steps_by_time_weekday), 
                         steps = mean_steps_by_time_weekday$steps_weekend, 
                         day = rep("Weekend", 288)))

# Need a label list to sent to lattice.

label_list <- list(at = seq(strptime("0000", format = "%H%M"), 
                            by = "6 hour", length = 5), 
                   labels = format(seq(strptime("0000", format = "%H%M"), 
                                       by = "6 hour", length = 5), "%H:%M"))

xyplot(steps ~ interval | day, data = mean_steps_by_time_weekday3, 
       layout = c(1,2), type = "l", scales = list(x = label_list))
```
